---
title: "Missing data, unreliable corr matrices, negative eigenvalues"
output: html_document
---


__Background__. I'd like to do a CCA. X will be imaging data. But, I have lots of missing values (result of unreliable white matter fiber estimation in small tracts).

__Set up__.

```{r setup,  warning=FALSE, message=FALSE, results='hide'}

#load libraries
libraries <- c('CCA', 'dplyr', 'matrixcalc') #list
lapply(libraries, require, character.only = T) #load 

#read in data
df <- read.csv(dir('../data/out/', full.names=T, pattern="^eigen")) 

```


__Data cleaning.__

```{r}

#identify variables with data from fewer than 10 participants 
(missing <- names(which(colSums(is.na(df)) > 10))) #n=29, all imaging vars are expected

#remove variables with data from fewer than 10 participants 
df<- df[, !colnames(df) %in% missing] 

```

__Create X and Y matrices.__
```{r}

#grep for relevant vars in df
FA   <- grep('FA', names(df), value=TRUE)
ncog <- grep('np_domain', names(df), value=TRUE)
scog <- grep('scog_', names(df), value=TRUE)
neg  <- grep('sans_sub_', names(df), value=TRUE)

#create X as imaging -- contains 66 vars
X <- scale(df[, names(df) %in% FA]) 

#create Y as ncog, scog, and neg -- contains 12 vars
Y <- c(ncog, scog, neg)
Y <- scale(df[, names(df) %in% Y]) 

```


__Calculate within and between correlation matrices__.
```{r}

#because of missing data, using pairwise complete observations
rxx <- cor(X, use = "p") ; ryy <- cor(Y, use = "p") ; rxy <- cor(X,Y, use = "p") 

#check if any _complete_ columns of correlation matrices are NA
sum(colSums(is.na(rxx)) > 0) ; sum(colSums(is.na(ryy)) > 0) ; sum(colSums(is.na(rxy)) > 0) #all 0

```

__But, omega calculation fails... (can't complete `chol`)__.
```{r}

#calculate omega
#omega = t(solve(chol(rxx))) %*% rxy %*% solve(chol(ryy)) #FAILS because can't compute chol(rxx)

#figure out why it fails - ensure symmetric
is.symmetric.matrix(rxx)
is.positive.definite(rxx) #uhoh

#figure out why it fails - ensure positive/non-0 eigenvalues
eigen(rxx)[1] #lots of negatives, and pretty big ones, too...

```

So, we see X is non-positive definite, and the negative eigenvalues (see last 14 values) are non-trivially large. 
This suggests that the correlation matrix estimation isn't robust, I think because I calculated pairwise correlation coefficients in the presence of missing values. So, I need to do something else about the missing data...

Unfortunately, N is not big enough to omit all observations with missing values (also I don't think it would be valid, given that so participants are expected to have some missing values, i.e., missing white matter tracts).

So, should I impute missing values? Or, should I "shrink" the correlation matrix (e.g., via the [Ledoit Wolf](http://www.ledoit.net/honey.pdf) method). Or something else?
